---
title: "MDD, BP"
output: html_document
date: "2025-07-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


QC
```{r}
# Install 
if (!requireNamespace("DESeq2")) install.packages("BiocManager"); BiocManager::install("DESeq2")
if (!requireNamespace("pheatmap")) install.packages("pheatmap")
if (!requireNamespace("RColorBrewer")) install.packages("RColorBrewer")
if (!requireNamespace("ggplot2")) install.packages("ggplot2")
if (!requireNamespace("vsn")) BiocManager::install("vsn")

# Load packages
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(ggplot2)
library(vsn)

# Create DESeqDataSet

metadata <- read.csv("ColData_Final.csv")
Count_Data <- read.csv("MDD_Final_Counts.csv", row.names = 1, header = TRUE, check.names = FALSE)

dds <- DESeqDataSetFromMatrix(countData = Count_Data,
                              colData = metadata,
                              design = ~ 1)  # no model yet, just QC

# Filter out low count genes
dds <- dds[rowSums(counts(dds)) > 1, ]

# Library size check
colSums(counts(dds))
barplot(colSums(counts(dds)), las=2, main="Library sizes", ylab="Total counts")

# Sample clustering and PCA
vsd <- varianceStabilizingTransformation(dds, blind=TRUE)  # variance-stabilizing transformation

# PCA
pcaData <- plotPCA(vsd, intgroup=colnames(metadata), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(x=PC1, y=PC2, color=metadata[[1]])) +
  geom_point(size=3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_bw()

# Sample distance heatmap
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- colnames(dds)
colnames(sampleDistMatrix) <- colnames(dds)
pheatmap(sampleDistMatrix, clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colorRampPalette(rev(brewer.pal(9, "Blues")))(255))

# Mean-variance relationship & Negative Binomial fit
dds_fit <- estimateSizeFactors(dds)
dds_fit <- estimateDispersions(dds_fit)
plotDispEsts(dds_fit)  # dispersion plot
# Red line: expected dispersion under the negative binomial
# Black points: gene-wise estimates
# Blue: final dispersion used

# Mean-sd plot (diagnosing heteroscedasticity) ===
meanSdPlot(assay(vsd))  # from vsn package

# Check for outliers using Cook's distance
dds_fit <- nbinomWaldTest(dds_fit)  # doesn't do testing, just needed to get mcols
cooks <- assays(dds_fit)[["cooks"]]
boxplot(log10(cooks + 1), las=2, main="Cook's distance (log10)")

# Top expressed genes
# Load required libraries
library(DESeq2)
library(tidyverse)
library(ggplot2)

# Use variance-stabilized data
vsd <- varianceStabilizingTransformation(dds, blind = TRUE)

# Get the normalized expression matrix
expr_matrix <- assay(vsd)

# Calculate mean expression for each miRNA
mean_expression <- rowMeans(expr_matrix)

# Create a data frame with miRNA names and mean expression
top_miRNAs_df <- data.frame(
  miRNA = rownames(expr_matrix),
  mean_expr = mean_expression
)

# Get top 20 most expressed miRNAs
top20 <- top_miRNAs_df %>%
  arrange(desc(mean_expr)) %>%
  slice_head(n = 20)

# Reorder
top20$miRNA <- factor(top20$miRNA, levels = top20$miRNA[order(top20$mean_expr)])

# Plot the top 20
ggplot(top20, aes(x = miRNA, y = mean_expr)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 20 Most Expressed miRNAs",
       x = "miRNA",
       y = "Mean Expression (VST normalized)") +
  theme_minimal(base_size = 14)

```


Outlier Identification
```{r}
# Using isotree
library(isotree)

dds <- estimateSizeFactors(dds)
idx <- rowSums( counts(dds, normalized=TRUE) >= 5 ) >= 3

# Running our model
dds <- dds[idx,]
dds <- DESeq(dds)
norm_counts <- counts(dds, normalized = TRUE)
sample_matrix <- t(norm_counts)

# Apply Isolation Forest
iso_model <- isolation.forest(sample_matrix, ntrees = 100)
outlier_scores <- predict(iso_model, sample_matrix, type = "score")

# Label outliers (e.g., top 5%)
threshold <- quantile(outlier_scores, 0.95)
outlier_samples <- which(outlier_scores >= threshold)

# View results
outlier_sample_names <- rownames(sample_matrix)[outlier_samples]

# Print names
outlier_sample_names

# got "352_E1" as > 0.5

# Cooks Distance
dds_fit <- estimateSizeFactors(dds)
dds_fit <- estimateDispersions(dds_fit)
# Check for outliers using Cook's distance
dds_fit <- nbinomWaldTest(dds_fit)  # doesn't do testing, just needed to get mcols
cooks <- assays(dds_fit)[["cooks"]]
boxplot(log10(cooks + 1), las=2, main="Cook's distance (log10)")

# PCA Plot
# Sample clustering and PCA
vsd <- varianceStabilizingTransformation(dds, blind=TRUE)  # variance-stabilizing transformation

# Attach sample metadata to PCA data
pcaData <- plotPCA(vsd, intgroup = colnames(metadata), returnData = TRUE)

# Add color group column from metadata (e.g., group by Stress, Run, etc.)
pcaData$Group <- metadata[[1]]  # or change this to metadata$Run or metadata$Stress

# Calculate variance from origin in PCA space
pcaData$PC_variance <- with(pcaData, PC1^2 + PC2^2)

# Select top 5 most variant samples
top10_samples <- top10_samples <- pcaData[order(-pcaData$PC_variance), ][1:10, ]

# Optional: add sample names if rownames are not informative
pcaData$Sample <- rownames(pcaData)
top10_samples$Sample <- rownames(top10_samples)

library(ggplot2)
library(ggrepel)

ggplot(pcaData, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 3) +
  geom_text_repel(
    data = top10_samples,
    aes(label = Sample),  # use Sample column, not rownames(top5_samples)
    size = 3.5,
    box.padding = 0.4
  ) +
  xlab(paste0("PC1: ", round(100 * attr(pcaData, "percentVar")[1]), "% variance")) +
  ylab(paste0("PC2: ", round(100 * attr(pcaData, "percentVar")[2]), "% variance")) +
  theme_bw()

# Cooks & PCA did not confirm 352_E1 as an global outlier, so all samples were kept for analysis.
```

Analysis
```{r}
# Analysis Packages
library( "DESeq2" )
library(ggplot2)
library(tidyverse)

# Load in data
metadata <- read.csv("ColData_Final_Anhodenia.csv")
Count_Data <- read.csv("MDD_Final_Counts.csv", row.names = 1, header = TRUE, check.names = FALSE)

# Make data a factor so it not treated as numerical (ID = numbers)
metadata$ID <- factor(metadata$ID)
metadata$Treatment <- factor(metadata$Treatment)

# Creating the deseq2 object with my design and running test
dds <- DESeqDataSetFromMatrix(countData = Count_Data,
                              colData = metadata,
                              design = ~ ID + Treatment)
dds <- estimateSizeFactors(dds)

# Removing genes that have low counts
idx <- rowSums( counts(dds, normalized=TRUE) >= 10 ) >= 3

# making pre-AAE the reference time point for the analysis.
dds$Treatment <- relevel(dds$Treatment, ref = "PRE")

dds <- dds[idx,]
dds <- DESeq(dds)

# Printing out how many genes failed to converge, along with the total. 
table(mcols(dds)$betaConv)

# Removal of genes that failed to converge into model (2%)
ddsClean <- dds[which(mcols(dds)$betaConv),]

# Constrasts and Results
results_75minPost_Pre <- results(ddsClean, contrast = c("Treatment","75 min POST","PRE"))
resFilt_75 <- results_75minPost_Pre[which(results_75minPost_Pre$padj < 0.10), ]
resnaomit_75 <- na.omit(resFilt_75)
write.csv(resnaomit_75, file="MDD_75minpost_7-15.csv")

results_Post_Pre <- results(ddsClean, contrast = c("Treatment","POST","PRE"))
resFilt_post <- results_Post_Pre[which(results_Post_Pre$padj < 0.10), ]
resnaomit_post <- na.omit(resFilt_post)
write.csv(resnaomit_post, file="MDD_post_7-15.csv")

results_Mid_Pre <- results(ddsClean, contrast = c("Treatment","MID","PRE"))
resFilt_mid <- results_Mid_Pre[which(results_Mid_Pre$padj < 0.10), ]
resnaomit_mid <- na.omit(resFilt_mid)
write.csv(resnaomit_mid, file="MDD_mid_7-15.csv")

# get n number per group
table(colData(dds)$Treatment)
# pre = 30, mid = 29, post = 28, and 75 min post = 28
```

PCA Plot (Figure 3a)
```{r}
vsd <- varianceStabilizingTransformation(ddsClean, blind = FALSE)
pca_data <- plotPCA(vsd, intgroup = "Treatment", returnData = TRUE)

# Define the comparisons manually and bind them
pre_mid  <- pca_data %>% filter(Treatment %in% c("PRE", "MID"))  %>% mutate(Facet = "PRE vs MID")
pre_post <- pca_data %>% filter(Treatment %in% c("PRE", "POST")) %>% mutate(Facet = "PRE vs POST")
pre_75   <- pca_data %>% filter(Treatment %in% c("PRE", "75 min POST")) %>% mutate(Facet = "PRE vs 75min POST")

# Combine them
subset_data <- bind_rows(pre_mid, pre_post, pre_75)

custom_colors <- c("PRE" = "red", "MID" = "darkgreen", "POST" = "navyblue", "75 min POST" = "purple")

ggplot(subset_data, aes(x = PC1, y = PC2, color = Treatment)) +
  geom_point(size = 3) +
  scale_color_manual(values = custom_colors) +
  facet_wrap(~Facet, scales = "free") +
  labs(title = "Principal Component Analysis of EV miRNA Expression: PRE vs. Post-Exercise Timepoints",
       x = paste0("PC1: ", round(attr(pca_data, "percentVar")[1] * 100), "% variance"),
       y = paste0("PC2: ", round(attr(pca_data, "percentVar")[2] * 100), "% variance")) +
  theme_minimal() +
  theme(strip.text = element_text(size = 12, face = "bold"),
        legend.position = "bottom")
```

Volcano Plots (Figure 3b)
```{r}
library(EnhancedVolcano)
resNAomit <- na.omit(results_75minPost_Pre)
res_df <- as.data.frame(resNAomit)

Sig <- res_df %>%
  filter(padj < 0.05) %>%
  rownames()

# Ensure log2FoldChange and padj columns are properly formatted
res_df$log2FoldChange <- as.numeric(res_df$log2FoldChange)
res_df$padj <- as.numeric(res_df$padj)

EnhancedVolcano(res_df,
                lab = rownames(res_df),   # Gene names (or set NULL if not needed)
                x = "log2FoldChange",     # X-axis: Log2 fold change
                y = "padj",               # Y-axis: Adjusted p-value
                title = "Post-AAE vs. Pre-AAE",
                subtitle = "Enduring Effect of AAE on EV miRNA Expression",
                selectLab = Sig,
                pCutoff = 0.05,           # P-value significance threshold
                FCcutoff = 0.1,             # Fold-change threshold (log2)
                pointSize = 1.0,          # Adjust point size
                labSize = 5.0,            # Label size
                colAlpha = 0.5,           # Transparency
                legendLabels = c("NS", "Log2FC", "P-value", "P-value + Log2FC"),
                legendPosition = "right",
                ylim = c(0,4), xlim = c(-4,6),
                drawConnectors = TRUE,
                col = c("gray", "blue", "green", "purple"),
                max.overlaps = 80
                )  # Colors: NS, Down, Up, Both
```

Heatmap of Shared DE miRNAs (Figure 4b)
```{r}
# Required libraries
library(DESeq2)
library(dplyr)
library(tidyr)
library(pheatmap)
library(tibble)

# Define miRNAs of interest
mirnas <- c("hsa-miR-3928-5p","hsa-miR-513a-3p", "hsa-miR-6792-5p", 
            "hsa-miR-20a-5p", "hsa-miR-5586-3p", "hsa-miR-504-5p", "hsa-miR-4710")

# Get normalized counts and subset to miRNAs
norm_counts <- counts(dds, normalized = TRUE)
mirna_counts <- norm_counts[rownames(norm_counts) %in% mirnas, ]

# Metadata with only desired timepoints
metadata <- as.data.frame(colData(dds)) %>%
  mutate(Sample = rownames(.)) %>%
  filter(Treatment %in% c("PRE", "MID", "POST", "75 min POST"))

# Reshape and join
mirna_long <- mirna_counts %>%
  as.data.frame() %>%
  rownames_to_column("miRNA") %>%
  pivot_longer(-miRNA, names_to = "Sample", values_to = "Expression") %>%
  left_join(metadata, by = "Sample") %>%
  filter(!is.na(Expression), !is.na(Treatment), !is.na(ID)) %>%
  mutate(log2Exp = log2(Expression + 1))

# Keep only IDs with all timepoints
valid_ids <- mirna_long %>%
  group_by(ID) %>%
  summarise(n_timepoints = n_distinct(Treatment), .groups = "drop") %>%
  filter(n_timepoints == 4)

mirna_clean <- mirna_long %>%
  semi_join(valid_ids, by = "ID")

# Pivot to compute log2FCs relative to PRE
fold_change <- mirna_clean %>%
  select(ID, miRNA, Treatment, log2Exp) %>%
  pivot_wider(names_from = Treatment, values_from = log2Exp) %>%
  mutate(
    MID_FC = MID - PRE,
    POST_FC = POST - PRE,
    `75minPOST_FC` = `75 min POST` - PRE
  ) %>%
  select(ID, miRNA, MID_FC, POST_FC, `75minPOST_FC`) %>%
  pivot_longer(cols = c(MID_FC, POST_FC, `75minPOST_FC`), 
               names_to = "Timepoint", values_to = "log2FC") %>%
  mutate(SampleID = paste(ID, Timepoint, sep = "_"))

# Create matrix for heatmap
heatmap_matrix <- fold_change %>%
  select(miRNA, SampleID, log2FC) %>%
  pivot_wider(names_from = SampleID, values_from = log2FC) %>%
  column_to_rownames("miRNA") %>%
  as.matrix()

# Order columns by MID → POST → 75minPOST
column_order <- fold_change %>%
  distinct(SampleID, Timepoint) %>%
  mutate(Timepoint = factor(Timepoint, levels = c("MID_FC", "POST_FC", "75minPOST_FC"))) %>%
  arrange(Timepoint) %>%
  pull(SampleID)

heatmap_matrix <- heatmap_matrix[, column_order]

# Order rows by average log2FC across timepoints (most upregulated at top)
row_order <- rowMeans(heatmap_matrix, na.rm = TRUE) %>%
  sort(decreasing = TRUE) %>%
  names()
heatmap_matrix <- heatmap_matrix[row_order, ]

# Color palette: red = downregulated, white = neutral, blue = upregulated
color_palette <- colorRampPalette(c("red", "white", "blue"))(100)

# Create annotation for columns
annotation_col <- data.frame(
  Timepoint = gsub(".*_(MID_FC|POST_FC|75minPOST_FC)", "\\1", colnames(heatmap_matrix))
)
annotation_col$Timepoint <- gsub("_FC", "", annotation_col$Timepoint)  # <- strip suffix
rownames(annotation_col) <- colnames(heatmap_matrix)

# Define custom colors
ann_colors <- list(
  Timepoint = c(
    "MID" = "darkgreen",
    "POST" = "lightblue",
    "75minPOST" = "purple"
  )
)

# Heatmap
pheatmap(
  heatmap_matrix,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = color_palette,
  na_col = "grey",
  main = "Log2 Fold Change vs PRE for Selected miRNAs",
  fontsize_row = 10,
  fontsize_col = 8,
  annotation_col = annotation_col,
  annotation_colors = ann_colors
)
```

Heatmap of all Sig miRNAs (Figure 4a)
```{r}
# Libraries
library(DESeq2)
library(dplyr)
library(tidyr)
library(tibble)
library(pheatmap)

# Get significant miRNAs for each timepoint (padj < 0.05)
get_sig <- function(res, label) {
  as.data.frame(res) %>%
    rownames_to_column("miRNA") %>%
    filter(padj < 0.05) %>%
    mutate(DE_timepoint = label)
}

# Get all significant DEGs
sig_mid  <- get_sig(results_Mid_Pre, "MID")
sig_post <- get_sig(results_Post_Pre, "POST")
sig_75   <- get_sig(results_75minPost_Pre, "75MIN")
sig_all  <- bind_rows(sig_mid, sig_post, sig_75)

# Determine DE pattern per miRNA
de_matrix <- sig_all %>%
  select(miRNA, DE_timepoint) %>%
  distinct() %>%
  mutate(present = 1) %>%
  pivot_wider(names_from = DE_timepoint, values_from = present, values_fill = 0) %>%
  mutate(
    DE_pattern = case_when(
      MID == 1 & POST == 0 & `75MIN` == 0 ~ "MID only",
      MID == 0 & POST == 1 & `75MIN` == 0 ~ "POST only",
      MID == 0 & POST == 0 & `75MIN` == 1 ~ "75MIN only",
      MID == 1 & POST == 1 & `75MIN` == 0 ~ "MID + POST",
      MID == 1 & POST == 0 & `75MIN` == 1 ~ "MID + 75MIN",
      MID == 0 & POST == 1 & `75MIN` == 1 ~ "POST + 75MIN",
      MID == 1 & POST == 1 & `75MIN` == 1 ~ "All timepoints",
      TRUE ~ "Other"
    )
  )

# Extract normalized counts
norm_counts <- counts(dds, normalized = TRUE) %>% as.data.frame()
norm_counts$miRNA <- rownames(norm_counts)

# Filter to only DE miRNAs
norm_counts <- norm_counts %>%
  filter(miRNA %in% de_matrix$miRNA)

# Melt to long format and join sample metadata
long_counts <- norm_counts %>%
  pivot_longer(-miRNA, names_to = "sample", values_to = "norm_count")

# Sample metadata 
meta <- as.data.frame(colData(dds)) %>%
  rownames_to_column("sample") %>%
  select(Sample_ID, ID, Treatment)

meta <- meta %>%
  rename("Sample_ID" = "sample")

long_counts <- long_counts %>%
  left_join(meta, by = "sample")

# Step 2: Average normalized counts per miRNA per subject per timepoint
long_avg <- long_counts %>%
  group_by(miRNA, ID, Treatment) %>%
  summarise(mean_count = mean(norm_count), .groups = "drop")


# Calculate per-subject paired log2FCs: (Treatment vs PRE)
# Pivot to wide so each row = one miRNA-subject; columns = PRE, MID, POST, 75MIN
wide_counts <- long_avg %>%
  pivot_wider(names_from = Treatment, values_from = mean_count)

# Log2FC calculations (pseudocount added)
get_log2fc <- function(df, timepoint) {
  df %>%
    filter(!is.na(PRE), !is.na(.data[[timepoint]])) %>%
    mutate(log2FC = log2((.data[[timepoint]] + 1) / (PRE + 1))) %>%
    select(miRNA, ID, log2FC) %>%
    mutate(Timepoint = timepoint)
}

fc_mid  <- get_log2fc(wide_counts, "MID")
fc_post <- get_log2fc(wide_counts, "POST")
fc_75   <- get_log2fc(wide_counts, "75 min POST")

log2fc_all <- bind_rows(fc_mid, fc_post, fc_75)

# Create column names: ID_Timepoint 
log2fc_all <- log2fc_all %>%
  mutate(sample_id = paste(ID, Timepoint, sep = "_")) %>%
  select(miRNA, sample_id, log2FC)

# Pivot to matrix format: rows = miRNA, columns = sample_id
lfc_matrix <- log2fc_all %>%
  pivot_wider(names_from = sample_id, values_from = log2FC) %>%
  column_to_rownames("miRNA")

# Order rows by DE pattern 
lfc_matrix <- lfc_matrix[match(de_matrix$miRNA, rownames(lfc_matrix)), ]

# Row and column annotations
# Row annotation (DE timepoint pattern)
row_anno <- data.frame(DE_Pattern = de_matrix$DE_pattern)
rownames(row_anno) <- de_matrix$miRNA

col_anno <- data.frame(sample_id = colnames(lfc_matrix)) %>%
  separate(sample_id, into = c("ID", "Timepoint"), sep = "_", remove = FALSE) %>%
  column_to_rownames("sample_id")

time_order <- c("MID", "POST", "75MIN")
ordered_cols <- rownames(col_anno)[order(match(col_anno$Timepoint, time_order))]
lfc_matrix <- lfc_matrix[, ordered_cols]
col_anno <- col_anno[ordered_cols, , drop = FALSE]

my_colors <- list(
  Timepoint = c(
    MID = "darkgreen",
    POST = "lightblue",
    `75 min POST` = "purple"
  )
)

# Plot heatmap 
pheatmap(
  lfc_matrix,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_row = row_anno,
  annotation_col = col_anno,
  annotation_colors = my_colors,
  color = colorRampPalette(c("red", "white", "blue"))(100),
  main = "Paired Log2FC (vs PRE) of DE miRNAs by Subject & Timepoint",
  show_colnames = FALSE,
  fontsize_row = 7,
  na_col = "grey90"
)
```

Getting fold change values (Based off of Figure 4b) (Figure 7b)
```{r}
# Keep only the relevant timepoints
mirna_sub <- mirna_long %>%
  filter(Treatment %in% c("PRE", "MID", "POST", "75 min POST"))

# log2 expression per timepoint
mirna_wide <- mirna_sub %>%
  select(ID, miRNA, Treatment, log2Exp) %>%
  pivot_wider(names_from = Treatment, values_from = log2Exp)

# Compute log2 fold changes 
mirna_fc <- mirna_wide %>%
  mutate(
    MID_FC        = ifelse(!is.na(MID)        & !is.na(PRE), MID        - PRE, NA),
    POST_FC       = ifelse(!is.na(POST)       & !is.na(PRE), POST       - PRE, NA),
    `75minPOST_FC`= ifelse(!is.na(`75 min POST`) & !is.na(PRE), `75 min POST` - PRE, NA)
  ) %>%
  select(ID, miRNA, MID_FC, POST_FC, `75minPOST_FC`)
```

Getting normalized counts 
```{r}
vsd <- vst(ddsClean, blind = FALSE)

# Extract VST expression matrix
vst_mat <- assay(vsd)

# Extract metadata
meta <- colData(vsd)

# Focus on one miRNA
mirna <- "hsa-miR-513a-3p"
mirna_vst <- vst_mat[mirna, ]
```

Bubble Plots for Pathway Analysis Results (Figure 6a)
```{r}
library(ggplot2)
library(readr)
library(dplyr)

# Load data
bubble_data <- read_csv("bubble_top10_all.csv")

# Convert pathway name to a factor, organized by timepoint, then significance
bubble_data$`GO Biological Process Complete Term` <- factor(bubble_data$`GO Biological Process Complete Term`, 
                           levels = rev(unique(bubble_data %>% arrange(`Time Point`, Bonferroni) %>% pull(`GO Biological Process Complete Term`))))


# Order timepoints 
bubble_data$`Time Point` <- factor(bubble_data$`Time Point`, levels = c("Mid-AAE", "Post-AAE", "75 minutes Post-AAE"))

# Plot bubble plot with fold enrichment instead of count
ggplot(bubble_data, aes(x = `Time Point`, y = `GO Biological Process Complete Term`)) +
  geom_point(aes(size = `Fold Enrichment`, color = -log10(Bonferroni))) +
  scale_color_viridis_c(option = "viridis", name = "-log10(Bonferroni)") +
  scale_size_continuous(name = "Fold Enrichment") +
  theme_minimal() +
  labs(
    title = "Top Enriched Pathways at Each Timepoint",
    x = "Time Point",
    y = "Enriched GO Term"
  ) +
  theme(axis.text.y = element_text(size = 9))

# Plot
ggplot(bubble_data, aes(x = `Time Point`, y = `GO Biological Process Complete Term`)) +
  geom_point(aes(size = Count, color = -log10(Bonferroni))) +
  scale_color_viridis_c(option = "viridis", name = "-log10(Bonferroni)") +
  scale_size_continuous(name = "Count") +
  theme_minimal() +
  labs(
    title = "Top Enriched Pathways at Each Timepoint",
    x = "Time Point",
    y = "Enriched GO Term"
  ) +
  theme(axis.text.y = element_text(size = 9))

```

Bubble Plot for 513a-3p Targets (Figure 7c)
```{r}
library(ggplot2)
library(viridis)

# Create the top 10 manually
top10_513a <- data.frame(
  GO_Term = c(
    "synapse assembly",
    "regulation of cellular response to transforming growth factor beta stimulus",
    "negative regulation of TOR signaling",
    "regulation of transforming growth factor beta receptor signaling pathway",
    "cellular response to leukemia inhibitory factor",
    "regulation of translational initiation",
    "negative regulation of mRNA metabolic process",
    "response to leukemia inhibitory factor",
    "miRNA processing",
    "dendrite morphogenesis"
  ),
  FoldEnrichment = c(2.44, 2.21, 2.69, 2.15, 2.31, 2.42, 2.29, 2.27, 3.39, 2.62),
  Bonferroni = c(0.0000571, 0.00188, 0.00307, 0.00566, 0.0186, 0.0219, 0.0229, 0.0284, 0.0300, 0.0358)
)

# Order GO terms for plotting
top10_513a$GO_Term <- factor(top10_513a$GO_Term, levels = rev(top10_513a$GO_Term))

# Plot
ggplot(top10_513a, aes(x = "", y = GO_Term)) +
  geom_point(aes(size = FoldEnrichment, color = -log10(Bonferroni))) +
  scale_color_viridis(option = "viridis", name = "-log10(Bonferroni)") +
  scale_size_continuous(name = "Fold Enrichment") +
  theme_minimal() +
  labs(
    title = "Top 10 Enriched Pathways for hsa-miR-513a-3p",
    x = NULL,
    y = "GO Biological Process"
  ) +
  theme(
    axis.text.y = element_text(size = 9),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank()
  )

```

Jaccard Test for Similarties & Differences in Target Genes (Figure 5b)
```{r}
jaccard <- function(a, b) {
  length(intersect(a, b)) / length(union(a, b))
}

gene_sets <- read_csv("gene_set_dendrite_morphogenesis.csv")

mid <- gene_sets$mid
post <- gene_sets$post
min75 <- gene_sets$min75

jaccard_matrix <- matrix(NA, 3, 3,
                         dimnames = list(c("mid", "post", "min75"),
                                         c("mid", "post", "min75")))

jaccard_matrix["mid", "post"] <- jaccard(mid, post)
jaccard_matrix["mid", "min75"] <- jaccard(mid, min75)
jaccard_matrix["post", "min75"] <- jaccard(post, min75)
jaccard_matrix[lower.tri(jaccard_matrix)] <- t(jaccard_matrix)[lower.tri(jaccard_matrix)]
diag(jaccard_matrix) <- 1

print(jaccard_matrix)

# How many genes are shared overall
mid_genes <- unique(na.omit(gene_sets$mid))
post_genes <- unique(na.omit(gene_sets$post))
min75_genes <- unique(na.omit(gene_sets$min75))

# mid vs post
shared_mid_post <- intersect(mid_genes, post_genes)
length(shared_mid_post)  # number of genes shared by mid and post

# post vs 75min
shared_post_75 <- intersect(post_genes, min75_genes)
length(shared_post_75)   # number of genes shared by post and 75min

# 75 minutes post-AAE vs mid-AAE
shared_75_mid <- intersect(min75_genes, mid_genes)
length(shared_75_mid)    # number of genes shared by 75min and mid

# View the shared genes
print(shared_mid_post)

pairwise_overlap <- data.frame(
  Comparison = c("mid vs post", "post vs 75min", "75min vs mid"),
  Shared_Genes = c(length(shared_mid_post), length(shared_post_75), length(shared_75_mid))
)

print(pairwise_overlap)

# Total Shared
shared_all_three <- Reduce(intersect, list(mid_genes, post_genes, min75_genes))
length(shared_all_three)  # Total number of shared genes

print(shared_all_three)

# Plots
gene_sets <- list(mid = mid_genes, post = post_genes, min75 = min75_genes)
jaccard <- function(a, b) {
  length(intersect(a, b)) / length(union(a, b))
}

# Initialize matrix
jaccard_matrix <- matrix(NA, nrow = 3, ncol = 3, 
                         dimnames = list(names(gene_sets), names(gene_sets)))

# Fill matrix
for (i in names(gene_sets)) {
  for (j in names(gene_sets)) {
    jaccard_matrix[i, j] <- jaccard(gene_sets[[i]], gene_sets[[j]])
  }
}

# Plot
pheatmap(jaccard_matrix,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         display_numbers = TRUE,
         main = "Jaccard Similarity Between Timepoints",
         color = colorRampPalette(c("white", "skyblue", "navy"))(100))

# Dendrite Morphogenesis, Shared Predicted Targets

gene_sets <- read_csv("gene_set_dendrite_morphogenesis.csv")

mid <- gene_sets$mid
post <- gene_sets$post
min75 <- gene_sets$'75min'

# How many genes are shared overall
mid_genes <- unique(na.omit(gene_sets$mid))
post_genes <- unique(na.omit(gene_sets$post))
min75_genes <- unique(na.omit(gene_sets$'75min'))

shared_all_three <- Reduce(intersect, list(mid_genes, post_genes, min75_genes))
length(shared_all_three)  # Total number of shared genes
```
