---
title: "Final Code MDD Project"
output: html_document
date: "2024-11-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Checking for outliers using OUTRIDER OUTlier 

```{r}

```


Creating the deseq object & running analysis

```{r cars}
# Open Necessary Libraries
library( "DESeq2" )
library(ggplot2)
library(tidyverse)

# Extract your ColData and CountData. Make sure that it in your working directory.
getwd() # get working directory
metadata <- read.csv('ColdataFinal.csv', header = TRUE, sep = ",") #extracting coldata
metadata$ID <- factor(metadata$ID)
metadata$Sex <- factor(metadata$Sex)
metadata$Treatment <- factor(metadata$Treatment)
Count_Data <- read.csv('Count_Data_MDD.csv', header = TRUE, sep = ",") # get count data, making sure that row headers in count data match sample ID. Nevermind the X in front of count data IDs.


# Create dds object to run analysis with appropriate model.
dds <- DESeqDataSetFromMatrix(countData = Count_Data,
                              colData = metadata,
                              design = ~ ID + Treatment, tidy = TRUE)

# Set your reference (control) group. 
dds$Condition <- relevel(dds$Treatment, ref = "PRE")

# Remove low counts from data before analysis
dds <- estimateSizeFactors(dds)
idx <- rowSums( counts(dds, normalized=TRUE) >= 5 ) >= 3
dds <- dds[idx,]

# Run analysis
dds <- DESeq(dds)
dds <- dds[which(mcols(dds)$betaConv),]

# Extract your results using a contrast
results_Post_vs_Pre <- results(dds, contrast = c("Treatment","POST","PRE"))
results_Mid_vs_Pre <- results(dds, contrast = c("Treatment","MID","PRE"))
results_75minPost_vs_Pre <- results(dds, contrast = c("Treatment","75 min POST","PRE"))

# Example of how to write a file with significant results. This will save to your directory.
resFilt <- results_Mid_vs_Pre[which(results_Mid_vs_Pre$padj < 0.10), ]
write.csv(resFilt, file="MDD_75minPost_PRE_Contrast.csv")
```

## Plots

Volcano Plot: Shows your upregulated/downregulated genes in one plot, and which onces have smaller p-values/greater fold change. 

```{r pressure, echo=FALSE}
res_volcano_plot <- lfcShrink(dds, contrast = c('Treatment','75 min POST','PRE'), results_75minPost_vs_Pre=results_75minPost_vs_Pre, type = 'normal') # prep data for plotting

EnhancedVolcano(res_volcano_plot, lab = rownames(res_volcano_plot), x = 'log2FoldChange', y = 'padj', pCutoff = 0.05, FCcutoff = 0.5) # plot DEGs that meet the 0.05 cut off, and show fold changes that are higher than 0.5.

plot # should show up under 'Plots'
```

PCA Plot for exploratory analysis - use this as a guide to determine which conditions would be interesting to look at. It is a PC2 vs. PC1 plot. 

```{r}
rld <- rlog(dds, blind=FALSE)
plot <- plotPCA(rld, intgroup = "Treatment", ntop = 500, returnData = FALSE) # uses log transformed data
```

